### 2015-05-14 background.tex
Space hacking.  Resorting to a bit of negative spacing.

\begin{figure}[t]
\begin{lstlisting}
static loff_t nvram_llseek(struct file *file,
    loff_t offset, int origin) {
  switch (origin) {
    case 0: break;
    case 1: offset += file->f_pos; break; // racy
    case 2: offset += nvram_len; break;
    default: offset = -1;
  }
  if (offset < 0)
    return -EINVAL;
  file->f_pos = offset; // racy
  return file->f_pos; // racy
}
\end{lstlisting}
\caption{Racy entry point in the generic\_nvram Linux driver}
\label{fig:data_race_example}
\end{figure}

The most common method for avoiding races is to protecting sets of program statements that access a shared resource using \emph{locks}, forming \emph{critical sections}.  
Figure~\ref{fig:lock_example} shows how to use a lock to eliminate the races in Figure~\ref{fig:data_race_example}. Because the return statement can potentially race on the \texttt{f\_pos} field, we store the result in a temporary variable inside the critical section.

Lock-based synchronization is a powerful tool for preventing races in drivers, but careless use of locks has many pitfalls~\cite{sutter2005software}. Coarse-grained locking can hurt performance, as it limits the opportunity for concurrency. On the other hand, fine-grained locking can easily lead to deadlocks. Although the Linux kernel provides sophisticated lock-free synchronization mechanisms~\cite{corbet2005linux}, such as atomic read-modify-write operations, in this work we focus on locks as they are widely used.\footnote{We treat lock-free operations soundly by not regarding them as providing any protection between threads, but this can lead to false alarm race reports.}

\begin{figure}[t]
\begin{lstlisting}
static loff_t nvram_llseek(struct file *file,
    loff_t offset, int origin) {
  mutex_lock(&nvram_mutex); // lock
  switch (origin) {
    case 0: break;
    case 1: offset += file->f_pos; break;
    case 2: offset += nvram_len; break;
    default: offset = -1;
  }
  if (offset < 0)
    return -EINVAL;
  file->f_pos = offset;
  loff_t res = file->f_pos; // store result
  mutex_unlock(&nvram_mutex); // unlock
  return res;
}
\end{lstlisting}
\caption{Using a lock to eliminate the data races in the previous example}
\label{fig:lock_example}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=1\linewidth]{img/lockset.pdf}
\caption{Applying lockset analysis on a thread of a concurrent program}
\label{fig:locksets}
\end{figure}

Lockset analysis for a concurrent program starts by creating a \emph{current} lockset, $\mathit{CLS}_T$, for each thread $T$ of the program, and a lockset, $\mathit{LS}_s$, for each shared variable $s$ used in the program. Initially, $\mathit{CLS}_T$ is empty for every thread $T$, because the threads do not hold any locks on program start. The lockset $\mathit{LS}_s$ for each variable $s$ is initialized to the set of all locks manipulated by the program, because initially each access to $s$ has (vacuously) been protected by every lock. The program is executed as usual (with threads scheduled according to the OS scheduler), except that instrumentation is added to update locksets as follows:
%
\begin{enumerate}
\item After each \emph{lock} and \emph{unlock} operation by $T$, $\mathit{CLS}_T$ is updated to reflect the locks currently held by $T$.
\item When $T$ accesses shared variable $s$, any locks that are not common to $\mathit{LS}_s$ and $\mathit{CLS}_T$ are removed from $\mathit{LS}_s$.
If $\mathit{LS}_s$ becomes empty as a result, a warning is issued that the access to $s$ may be unprotected.
\end{enumerate}

@@

>>>>

\begin{figure}[t]
\begin{lstlisting}
static loff_t nvram_llseek(struct file *file,
    loff_t offset, int origin) {
  switch (origin) {
    case 0: break;
    case 1: offset += file->f_pos; break; // racy
    case 2: offset += nvram_len; break;
    default: offset = -1;
  }
  if (offset < 0)
    return -EINVAL;
  file->f_pos = offset; // racy
  return file->f_pos; // racy
}
\end{lstlisting}
\vspace{-2mm}
\caption{Racy entry point in the generic\_nvram Linux driver}
\label{fig:data_race_example}
\end{figure}

The most common method for avoiding races is to protecting sets of program statements that access a shared resource using \emph{locks}, forming \emph{critical sections}.  
Figure~\ref{fig:lock_example} shows how to use a lock to eliminate the races in Figure~\ref{fig:data_race_example}. Because the return statement can potentially race on the \texttt{f\_pos} field, we store the result in a temporary variable inside the critical section.
%
Careless use of locks has many pitfalls~\cite{sutter2005software}: coarse-grained locking can hurt performance, as it limits the opportunity for concurrency, while fine-grained locking can easily lead to deadlocks. Although the Linux kernel provides sophisticated lock-free synchronization mechanisms~\cite{corbet2005linux}, such as atomic read-modify-write operations, in this work we focus on locks as they are widely used.\footnote{We treat lock-free operations soundly by not regarding them as providing any protection between threads, but this can lead to false alarm race reports.}

\begin{figure}[t]
\begin{lstlisting}
static loff_t nvram_llseek(struct file *file,
    loff_t offset, int origin) {
  mutex_lock(&nvram_mutex); // lock
  switch (origin) {
    case 0: break;
    case 1: offset += file->f_pos; break;
    case 2: offset += nvram_len; break;
    default: offset = -1;
  }
  if (offset < 0)
    return -EINVAL;
  file->f_pos = offset;
  loff_t res = file->f_pos; // store result
  mutex_unlock(&nvram_mutex); // unlock
  return res;
}
\end{lstlisting}
\vspace{-2mm}
\caption{Using a lock to eliminate the data races in the previous example}
\label{fig:lock_example}
\end{figure}

@@

Lockset analysis for a concurrent program starts by creating a \emph{current} lockset, $\mathit{CLS}_T$, for each thread $T$ of the program, and a lockset, $\mathit{LS}_s$, for each shared variable $s$ used in the program. Initially, $\mathit{CLS}_T$ is empty for every thread $T$, because the threads do not hold any locks on program start. The lockset $\mathit{LS}_s$ for each variable $s$ is initialized to the set of all locks manipulated by the program, because initially each access to $s$ has (vacuously) been protected by every lock. The program is executed as usual (with threads scheduled according to the OS scheduler), except that instrumentation is added to update locksets as follows.
%
After each \emph{lock} and \emph{unlock} operation by $T$, $\mathit{CLS}_T$ is updated to reflect the locks currently held by $T$.
%
When $T$ accesses shared variable $s$, any locks that are not common to $\mathit{LS}_s$ and $\mathit{CLS}_T$ are removed from $\mathit{LS}_s$.
If $\mathit{LS}_s$ becomes empty as a result, a warning is issued that the access to $s$ may be unprotected.

@@

\begin{figure}[t]
\centering
\includegraphics[width=1\linewidth]{img/lockset.pdf}
\caption{Applying lockset analysis on a thread of a concurrent program}
\label{fig:locksets}
\end{figure}

### 2015-05-14 corral.tex
Space hacking.  Resorting to a bit of negative spacing.

We have implemented two different levels of context-switch instrumentation granularity:

\yieldcoarse instruments \texttt{yield} statement in a binary fashion: if \whoop
finds that a pair of entry points cannot race, then it will only instrument
context-switches in synchronization points (i.e.\ locks and unlocks) of this pair; else if \whoop finds
that a pair of entry points can race, then it will instrument context-switches
in all visible operations of this pair.

\yieldmr is a finer-grained instrumentation: \whoop instruments context-switches
in all synchronization points, but only instruments context-switches in memory
regions that can potentially race (regardless if a pair has not been fully
verified as race-free). In our experiments (see \S\ref{evaluation}),
\yieldmr performs significantly faster than \yieldcoarse.

>>>>

We have implemented two different levels of context-switch instrumentation granularity.
%
\textbf{\yieldcoarse} instruments \texttt{yield} statement in a binary fashion: if \whoop
finds that a pair of entry points cannot race, then it will only instrument
context-switches in synchronization points (i.e.\ locks and unlocks) of this pair; else if \whoop finds
that a pair of entry points can race, then it will instrument context-switches
in all visible operations of this pair.
%
\textbf{\yieldmr} is a finer-grained instrumentation: \whoop instruments context-switches
in all synchronization points, but only instruments context-switches in memory
regions that can potentially race (regardless if a pair has not been fully
verified as race-free). In our experiments (see \S\ref{evaluation}),
\yieldmr significantly outperforms \yieldcoarse.

### 2015-05-14 evaluation.tex
Space hacking.  Resorting to a bit of negative spacing.

Using the race-freedom guarantees from \whoop, we managed to significantly accelerate \corral in most of our benchmarks; the best results were achieved using \yieldmr. Figure~\ref{fig:plot} shows that most speedups using \yieldmr are in the range of 1.5--10$\times$; while in ssu100, and pc8736x\_gpio respectively, we achieved a speedup of 12$\times$, and 20$\times$ respectively, using a csb of 9. We notice that a higher csb typically results into greater speedups, when exploiting \whoop. This is expected, as a higher csb directly translates into a larger sequentialized program, and thus \whoop has more opportunities to reduce the sequentialization.

However, if \whoop fails to verify any race-related properties of a driver, it might actually slow down \corral. We noticed this in the sx8 driver, where \whoop does not verify any pairs, causing unnecessary overhead. Although \whoop verified 46 out of 47 memory regions, it still fails to speedup \corral using \yieldmr. This might be because \corral verifies the only 2 pairs of sx8 quite fast (21.4 seconds with csb of 9), and thus the overhead of using \whoop overwhelms any benefits from our fine-grained instrumentation.

>>>>

Using the race-freedom guarantees from \whoop, we managed to significantly accelerate \corral in most of our benchmarks; the best results were achieved using \yieldmr. Figure~\ref{fig:plot} shows that most speedups using \yieldmr are in the range of 1.5--10$\times$; while in ssu100, and pc8736x\_gpio respectively, we achieved a speedup of 12$\times$, and 20$\times$ respectively, using a csb of 9. We notice that a higher csb typically results into greater speedups, when exploiting \whoop. This is expected, as a higher csb directly translates into a larger sequentialized program, and thus \whoop has more opportunities to reduce the sequentialization.
%
However, if \whoop fails to verify any race-related properties of a driver, it can slow down \corral. We noticed this in the sx8 driver, where \whoop does not verify any pairs, causing unnecessary overhead. Although \whoop verified 46 out of 47 memory regions, it still fails to speedup \corral using \yieldmr. \corral checks the only two pairs associated with sx8 quite quickly (21.4 seconds with csb of 9), so that the overhead of running \whoop outweighs the benefits of fine-grained instrumentation.

### 2015-05-14 intro.tex
Space hacking.  Resorting to a bit of negative spacing.

A major reason behind these concurrency bugs is that drivers operate inside a complicated and hostile environment~\cite{corbet2005linux}. For example the Linux kernel can invoke its drivers from multiple concurrent threads, a hardware device can issue interrupt requests that cause running processes to block and switch execution context, and the user may decide to remove a device (hot-plugging) while some operation is still running. All these scenarios can easily cause \emph{data races} if no proper synchronization mechanisms (e.g.\ locks) are in place to control concurrent access to shared resources.

Data races are a source of undefined behavior in C, and lead to nondeterministically occurring bugs that can be hard to reproduce, isolate and fix, especially in the context of complex operating systems.
%
Several techniques have been successfully used to analyze device drivers~\cite{ball2006thorough, clarke2004predicate, engler2000checking, henzinger2002temporal, cook2006termination, kuznetsov2010testing, renzelmann2012symdrive, lal2012corral}, but most focus on generic sequential program properties and protocol bugs. Linux kernel analyzers, such as sparse~\cite{corbet2004sparse}, coccinelle~\cite{padioleau2008doc} and lockdep~\cite{corbet2006lock}, can find deadlocks in kernel source code, but are unable to detect races. Techniques that can detect races in drivers~\cite{qadeer2004kiss, pratikakis2006locksmith, voung2007relay, lal2012corral} are usually either \emph{unsound} (i.e.\ can miss real bugs) or \emph{imprecise} (i.e.\ can report false bugs), and typically sacrifice precision for scalability. Thus, there is a clear need for new tools that are able to detect races efficiently and precisely.

In this paper, we present \whoop, a new fully automatic approach that statically analyzes device drivers written in C for data races. \whoop is empowered by \emph{symbolic pairwise lockset analysis}, a novel technique for verifying the absence of data races in device drivers. We attempt to prove a driver race-free by: (i) deriving a sound \emph{sequential} program that \emph{over-approximates} the originally concurrent program; (ii) instrumenting it to record \emph{locksets}; and (iii) using the locksets to assert that all accesses to the same shared resource are \emph{consistently} protected by at least one common lock. Reducing analysis to reasoning over a sequential program avoids the need to enumerate thread interleavings, and allows reuse of existing successful sequential verification techniques.

We then show that by exploiting the race-freedom guarantees provided by our symbolic analysis we can achieve a sound partial-order reduction that significantly accelerates \corral~\cite{lal2012corral}, a state-of-the-art precise bug-finder used by Microsoft to analyze drivers, as well as other concurrent programs. Using \whoop and \corral we analyzed \sizeOfBenchmarks drivers from the 4.0 Linux kernel distribution.  We find that by combining \whoop and \corral we achieve concurrency analysis speedups in the range of 1.5--10$\times$ for most of our benchmarks, compared with using \corral in isolation.  For two drivers, we observe even greater speedups of 12$\times$ and 20$\times$.

\whoop currently supports Linux drivers, but our approach is conceptually generic, thus could be applied to analyze drivers for other operating systems (e.g.\ Windows) as well as other concurrent systems that use a similar programming model (e.g.\ filesystems).

>>>>

A major cause of concurrency bugs is the complicated and hostile environment in which drivers operate~\cite{corbet2005linux}. The Linux kernel can invoke drivers from multiple threads, a hardware device can issue interrupt requests that cause running processes to block and switch execution context, and the user may decide to remove a device (hot-plugging) while some operation is still running. All these scenarios can easily cause \emph{data races} if no proper synchronization mechanisms (e.g.\ locks) are in place to control concurrent access to shared resources.
%
Data races are a source of undefined behavior in C, and lead to nondeterministically occurring bugs that can be hard to reproduce, isolate and fix, especially in the context of complex operating systems.
%
Several techniques have been successfully used to analyze device drivers~\cite{ball2006thorough, clarke2004predicate, engler2000checking, henzinger2002temporal, cook2006termination, kuznetsov2010testing, renzelmann2012symdrive, lal2012corral}, but most focus on generic sequential program properties and protocol bugs. Linux kernel analyzers, such as sparse~\cite{corbet2004sparse}, coccinelle~\cite{padioleau2008doc} and lockdep~\cite{corbet2006lock}, can find deadlocks in kernel source code, but are unable to detect races. Techniques that can detect races in drivers~\cite{qadeer2004kiss, pratikakis2006locksmith, voung2007relay, lal2012corral} are usually either \emph{unsound} (i.e.\ can miss real bugs) or \emph{imprecise} (i.e.\ can report false bugs), and typically sacrifice precision for scalability. Thus, there is a clear need for new tools that are able to detect races efficiently and precisely.

In this paper, we present \whoop, a new fully automatic approach that statically analyzes device drivers written in C for data races. \whoop is empowered by \emph{symbolic pairwise lockset analysis}, a novel technique for verifying the absence of data races in device drivers. We attempt to prove a driver race-free by: (i) deriving a sound \emph{sequential} program that \emph{over-approximates} the originally concurrent program; (ii) instrumenting it to record \emph{locksets}; and (iii) using the locksets to assert that all accesses to the same shared resource are \emph{consistently} protected by at least one common lock. Reducing analysis to reasoning over a sequential program avoids the need to enumerate thread interleavings, and allows reuse of existing successful sequential verification techniques.
%
We then show that by exploiting the race-freedom guarantees provided by our symbolic analysis we can achieve a sound partial-order reduction that significantly accelerates \corral~\cite{lal2012corral}, a state-of-the-art precise bug-finder used by Microsoft to analyze drivers, as well as other concurrent programs. Using \whoop and \corral we analyzed \sizeOfBenchmarks drivers from the 4.0 Linux kernel distribution.  We find that by combining \whoop and \corral we achieve concurrency analysis speedups in the range of 1.5--10$\times$ for most of our benchmarks, compared with using \corral in isolation.  For two drivers, we observe even greater speedups of 12$\times$ and 20$\times$.

\whoop currently supports Linux drivers, but our approach is conceptually generic and could be applied to analyze drivers for other operating systems as well as to concurrent systems that use a similar programming model (e.g.\ filesystems).

### 2015-05-14 whoop.tex
Space hacking.  Resorting to a bit of negative spacing.

We now present \emph{symbolic pairwise lockset analysis}, a novel method for data race analysis in device drivers.  In \S\ref{sec:symbolicpairwise} we describe how the approach works in a semi-formal manner, with respect to a simple concurrent programming model.  In \S\ref{sec:implementation} we explain how we have implemented symbolic pairwise lockset analysis in a practical tool, \whoop, that can be applied directly to driver source code.

Our experimental evaluation (\S\ref{evaluation}) demonstrates that the \whoop technique has value as a stand-alone analyzer, and that results from \whoop can be exploited to significantly boost the performance of a more precise symbolic analysis for concurrency, based on context bounding and stratified inlining, offered by the \corral~\cite{lal2012corral} tool; we discuss the latter approach in \S\ref{corral}.

@@

At the start of execution the valuation of shared variables is arbitrary, no mutexes are held (i.e.\ each mutex maps to $\bot$), and the list of threads is empty.
%
At any point of execution, a new thread may be added to the list of threads.  This involves selecting a thread template $T$ and an id $i \in \ids$ that has not been previously used during program execution, setting the point of execution for the new thread to be the first statement of $\mainforthread{T}$, and choosing an arbitrary valuation for the private variables $\variables{T}$.

We consider a standard interleaving model of concurrency: at any execution point, a thread may execute its current statement, unless that statement has the form $\mutexlock{m}$ and mutex $m$ is already held by some thread.  Executing a statement causes the state of the thread, and the shared state, to be updated in a standard manner.  For example, if a thread following template $T$ executes $s = e$, where $s \in \sharedvars$ and $e$ is an expression over $\variables{T}$, the shared variable valuation is updated so that $s$ has the value determined by evaluating $e$ in the context of the thread's private variable valuation.  Because our interest is in data race analysis for race-free programming, we are not concerned with relaxed memory behavior: race-free programs exhibit only sequentially consistent behaviors.

@@

Lockset instrumentation introduces: sets $\hasbeenread{i} \subseteq \powerset{\sharedvars}$ and $\hasbeenwritten{i} \subseteq \powerset{\sharedvars}$ to track the shared variables that have been read from and written to, respectively, by the thread executing $T$; a current lockset, $\currentlockset{i} \subseteq \powerset{\mutexes}$, to record the mutexes that are currently held by the thread; and, for each shared variable $s \in \sharedvars$, a lockset, $\lockset{s}{i}$, to record the mutexes that are consistently held when the thread accesses $s$.

The statements of each procedure in $\instrument{T}{i}$ that access shared variables and mutexes are instrumented to manipulate these sets.  This transformation is described in Figure~\ref{fig:instrumentation}.  A shared variable assignment, $s = e$, is instrumented by recording in $\hasbeenwritten{i}$ that $s$ has been written to, and updating the lockset for $s$, $\lockset{s}{i}$, to eliminate any mutexes that are not currently held (i.e.\ those mutexes that are not in $\currentlockset{i}$).  A shared variable read, $x = s$ operates analogously, except for the additional $\havockeyword$ command which we discuss under \emph{Shared State Abstraction} below.  Instrumentation of mutex manipulation commands, $\mutexlock{m}$ and $\mutexunlock{m}$, involves updating the current lockset, $\currentlockset{i}$, to add and remove mutex $m$, respectively.

@@

We require the threads to terminate because in the presence of non-termination the assertion at the end of Figure~\ref{fig:sequentialization} may not be reached.  The termination analysis problem for device drivers has been widely studied, notably in the context of the Terminator tool~\cite{cook2006termination}. In the remainder of the paper we do not consider termination issues, assuming that the drivers we analyze in our experimental evaluation (see \S\ref{evaluation}) are indeed terminating.

>>>>

We now present \emph{symbolic pairwise lockset analysis}, a novel method for data race analysis in device drivers.  In \S\ref{sec:symbolicpairwise} we describe how the approach works in a semi-formal manner, with respect to a simple concurrent programming model.  In \S\ref{sec:implementation} we explain how we have implemented symbolic pairwise lockset analysis in a practical tool, \whoop, that can be applied directly to driver source code.
%
Our experimental evaluation (\S\ref{evaluation}) demonstrates that the \whoop technique has value as a stand-alone analyzer, and that results from \whoop can be exploited to significantly boost the performance of a more precise symbolic analysis for concurrency, based on context bounding and stratified inlining, offered by the \corral~\cite{lal2012corral} tool; we discuss the latter approach in \S\ref{corral}.

@@

At the start of execution the valuation of shared variables is arbitrary, no mutexes are held (i.e.\ each mutex maps to $\bot$), and the list of threads is empty.
%
At any point of execution, a new thread may be added to the list of threads.  This involves selecting a thread template $T$ and an id $i \in \ids$ that has not been previously used during program execution, setting the point of execution for the new thread to be the first statement of $\mainforthread{T}$, and choosing an arbitrary valuation for the private variables $\variables{T}$.
%
We consider a standard interleaving model of concurrency: at any execution point, a thread may execute its current statement, unless that statement has the form $\mutexlock{m}$ and mutex $m$ is already held by some thread.  Executing a statement causes the state of the thread, and the shared state, to be updated in a standard manner.  For example, if a thread following template $T$ executes $s = e$, where $s \in \sharedvars$ and $e$ is an expression over $\variables{T}$, the shared variable valuation is updated so that $s$ has the value determined by evaluating $e$ in the context of the thread's private variable valuation.  Because our interest is in data race analysis for race-free programming, we are not concerned with relaxed memory behavior: race-free programs exhibit only sequentially consistent behaviors.

@@

Lockset instrumentation introduces: sets $\hasbeenread{i} \subseteq \powerset{\sharedvars}$ and $\hasbeenwritten{i} \subseteq \powerset{\sharedvars}$ to track the shared variables that have been read from and written to, respectively, by the thread executing $T$; a current lockset, $\currentlockset{i} \subseteq \powerset{\mutexes}$, to record the mutexes that are currently held by the thread; and, for each shared variable $s \in \sharedvars$, a lockset, $\lockset{s}{i}$, to record the mutexes that are consistently held when the thread accesses $s$.
%
The statements of each procedure in $\instrument{T}{i}$ that access shared variables and mutexes are instrumented to manipulate these sets, as shown in Figure~\ref{fig:instrumentation}.  For a shared variable assignment $s = e$ we record in $\hasbeenwritten{i}$ that $s$ has been written to, and update the lockset $\lockset{s}{i}$ to eliminate any mutexes that are not currently held (those mutexes that are not in $\currentlockset{i}$).  A shared variable read $x = s$ is instrumented analogously, with an additional $\havockeyword$ command which we discuss under \emph{Shared State Abstraction} below.  Instrumentation of mutex manipulation commands, $\mutexlock{m}$ and $\mutexunlock{m}$, involves updating the current lockset, $\currentlockset{i}$, to add and remove mutex $m$, respectively.

@@

In the presence of non-termination the assertion at the end of Figure~\ref{fig:sequentialization} may not be reached.  The termination analysis problem for device drivers has been widely studied (see e.g.\ ~\cite{cook2006termination}), and in the remainder of the paper we do not consider termination issues, assuming that the drivers we analyze in our experimental evaluation (see \S\ref{evaluation}) are terminating.
